---
title: "Getting started"
output: rmarkdown::html_vignette
bibliography: vignette.bib
vignette: >
  %\VignetteIndexEntry{ht}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sapo)
library(sf)

set.seed(2024)
```

> The `sapo` package provides a Monte Carlo test for evaluating spatial
> association between polygons. For reproducibility, set a seed before using the
> functions.


This vignette illustrates the functionality of the `sapo` package. `sapo`, which
is Portuguese for "toad," stands for "**S**patial **A**ssociation Between
**Po**lygons".  This package enables you to test whether two types of polygons
are spatially associated. This raises two key questions: 1) What do these
"polygon types" represent? and 2) What constitutes spatial association?


Polygons can represent various spatial entities. In ecology, for instance,
polygons might represent ecological patches [@mcgarigal2013landscape], animal
home ranges, or forest canopy gaps. It's often important to determine if two
types of spatial phenomena (like those mentioned above) tend to attract or repel
each other.  Conditional Monte Carlo (MC) hypothesis tests [@godoy2022testing]
offer a way to test these hypotheses. The `cmc_psat` function (conditional MC
polygon association test) implements these methods. All methods are based on
conditional MC simulation, generating spatial patterns that assume spatial
independence (using an adaptation of the toroidal shift^[a technique that shifts
the polygons to simulate different spatial arrangements]) while preserving the
observed marginal spatial structures.


To illustrate the package's functionality, consider two simulated polygon types,
`poly1` and `poly2`, loaded below as `sf` objects. These will represent two
distinct spatial features. Our goal is to test whether they are _spatially
independent_.

```{r}
poly1 <- system.file("extdata", "poly1.rds", package = "sapo") |>
  readRDS()
poly2 <- system.file("extdata", "poly2.rds", package = "sapo") |>
  readRDS()

str(poly1)
str(poly2)
```

Below, we plot these polygon types from this toy example.

```{r}
plot(sf::st_geometry(poly1), col = "gray70")
plot(sf::st_geometry(poly2), add = TRUE)
```

The `cmc_psat` function requires that its first two arguments, `p1` and `p2`,
are `sf` objects. Other important arguments include:

*   `n_sim` (default = 499): An integer specifying the number of conditional MC
    simulations under the null hypothesis of spatial independence.

*   `alpha` (default = 0.05): A real number (between 0 and 1) indicating the
    significance level for the hypothesis test.


For a deeper understanding of the remaining arguments, refer to
@godoy2022testing, which provides a comprehensive explanation of the method. The
default argument values generally optimize the test's power and type I error
rate.


The code below tests whether `poly1` and `poly2` are _spatially independent_. In
essence, it assesses whether the presence of a type 2 polygon affects the
likelihood of observing a type 1 polygon.

```{r}
my_ht <- cmc_psat(poly1, poly2)
```

The test's p-value is `r sprintf("%.4f", my_ht$p.value)`, indicating a lack of
spatial dependence. If the analysis suggests that the polygons are not spatially
independent, a descriptive analysis can be performed using a generalization of
the $H_{12}$ function (typically $K_{12}$ or $L_{12}$, with the latter being the
package default) used for analyzing spatial point patterns.


We'll now create point-wise confidence bands based on the $L_{12}$ null
hypothesis. These bands will be plotted against the $L_{12}$ function calculated
from the real data. A significant result with the black curve (representing the
real data) below the envelope suggests that the two polygon types repel each
other. Conversely, if the black curve is above the envelope, it indicates
attraction. Remember that these envelopes are for descriptive analysis, not
inference. The _p-value_ determines whether to reject the hypothesis of spatial
independence.
```{r}
sig_level <- my_ht$alpha
sig_plot <-
  data.frame(r = my_ht$distances,
             h = my_ht$mc_funct[NROW(my_ht$mc_funct), ],
             l = apply(my_ht$mc_funct, 2, quantile,
                       prob = .5 * sig_level),
             u = apply(my_ht$mc_funct, 2, quantile,
                       prob = 1 - .5 * sig_level))
## par(mfrow = c(1, 2))
## ## plot density part
## den <- density(my_ht$mc_sample)
## plot(den, main = sprintf("Test statistic (p-value = %.4f)",
##                          my_ht$p_value),
##      xlab = "u", ylab = "density")
## value <- my_ht$mc_sample[length(my_ht$mc_sample)]
## polygon(c(den$x[den$x >= value], value),
##         c(den$y[den$x >= value], 0),
##         col = "coral1",
##         border = "transparent")
## lines(den)
## H function
with(sig_plot, plot(x = r, y = h, type = "l",
                    col = 2,
                    main = "Local envelope",
                    xlab = "r",
                    ylab = expression(H[12](r)),
                    ylim = range(c(h, l, u))))
with(sig_plot, lines(x = r, y = l, lty = 2, col = 2))
with(sig_plot, lines(x = r, y = u, lty = 2, col = 2))
with(sig_plot,
     polygon(c(r, rev(r)), c(l, rev(u)),
             col = "coral1", lty = 0))
with(sig_plot, lines(x = r, y = h,
                     col = 1))
```
